Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-07 3.Thesis/Writeup/2.Manuscript.Rmd
After feedback Gerko (not yet with notes meeting)

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence of iterative imputation algorithms."
keywords: missing data, iterative imputation, non-convergence, mice
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
---

@@

@@

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous, but treating it can also be tedious. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to *impute* (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances [@buur18].

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. Missing data $y_{mis}$ in an incomplete dataset is imputed $M$ times. The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets. On each completed dataset, an analysis of scientific interest is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained. These $M$ values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but missing---data inference.

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values $y_{imp}$ are drawn from the posterior predictive distribution of the missing values $y_{mis}$. The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. That is, the state space of the algorithm is monitored by plotting a scalar summary $\theta$ against the iteration number.
**[Segway to WHAT TO MONITOR? Also, this paragraph is more or less the same as the third paragraph in the second section. Merge or remove one of the two?]** Typically, the $\theta$s under evaluation are chain means and variances (**define!**). These $\theta$s may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of $y_{imp}$. A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient). Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. Therefore, this $\theta$ is not sufficient either. Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented. 
<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
**[Segway to HOW TO MONITOR?]** Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2]. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics  may only identify signs of *non*-convergence [@hoff09]. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{mis}$. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] **

# Identifying non-convergence

@@

**[Move to INTRO???]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$. We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $y_{imp,m}$, plotted against the iteration number. Each line portrays one imputation. The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. **[Leave this in if it's after mixing and stationarity are defined, otherwise remove?] In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of $j$ in $y_{imp, m}$. This difference (presumably) translates to bias in the $M$ completed data estimates $\hat{\mu}_{j,m}$, and consequently also to the pooled estimate $\bar{\mu}_j$. Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

**[Move to INTRO???]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain $\theta$. $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. As @buur18 describes, researchers can specify their quantity of scientific interest $Q$ (e.g., a regression coefficient) as scalar summary $\theta$. Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.  The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a $\theta$ that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose $\lambda_1$ as such a scalar summary. We define $\lambda_1$ as the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$. $\lambda_1$ is measure that summarizes the covariances in the completed datasets. **The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by @mack03.**

@@

## Diagnostics under evaluation

@@

**[Add: effective sample size is used if we want independent samples *within* chains. In MI, we only use the final estimate. Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand. But the default in iterative imp is often substantially lower.]**

### Potential scale reduction factor

@@

### Autocorrelation

@@

### Thresholds

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$. Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**. And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice. That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$. **[and formulate advice on thetas?]**

@@

@@

5. **[Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas. The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.]**

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14]. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem **(from an unspecified social scientific field of study)**. Let the predictor space be defined as 
$$
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
$$
A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome $Y$. For each unit $i = 1, 2,..., N$, let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s. **[MOVE this next part to the start of the Simulation Study section?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$. We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**

@@

We consider four types of $Q$s that are often of interest in empirical research. Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j = Y, X_1, X_2, X_3$. We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
where $Y'$ is the expected value of the outcome. And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2 \times 100$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**. The true values of these scientific estimands are calculated on the complete data with the `R` functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R]. 
<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).  -->

@@

With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the $\bar{Q}$s are *not* unbiased, confidence valid estimates of the $Q$s. **[Rephrase this:] We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.**

**[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s. More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$. Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$). We look at averages across simulation repetitions. Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergence diagnostics $\widehat{R}$ and $AC$. We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.**

@@

**[Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) Add more info about figure legends and axes]**

@@

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s. **[Add panel labels, describe panels]** Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$. The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$. The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$. This implies that, under the current specifications, no more than seven iterations are necessary to obtain *sufficient* performance in terms of bias and confidence validity. **[only substantial effect of non-conv in multivariate $Q$s]**

### $Q=\mu$.
The estimated univariate means seem unaffected by the number of iterations in the algorithm. This implies that *suffient* inferences may be obtained with as little as one iteration ($T \geq 1$). The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% causing more extreme bias than other conditions. Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$. Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is relatively small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.

>>>>

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence of iterative imputation algorithms. We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity."
keywords: missing data, iterative imputation, non-convergence, mice
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
---

@@

# Overview of sections

Intro

- This is missingness

- This is how we solve missingness

- For that we need an algorithm, but what if it has not converged?

- That is what we solve in this paper

Algorithmic convergence

- What is algorithmic convergence?

- When is it converged? Stationarity versus mixing.

- How is convergence currently investigated? And how should it be investigated? Multivariate versus univariate.

Methods

- Which methods are there? And how do they relate to mixing and stationarity?

- $\widehat{R}$

- AC

- We can use these on any $\theta$

The Simulation study, Results, Discussion.

@@

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to *impute* (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances [@buur18].

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part $y_{mis}$ of an incomplete dataset is imputed $M$ times. The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets. On each completed dataset, an analysis of scientific interest is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained. These $M$ values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but unobserved---inference.

@@

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values $y_{imp}$ are drawn from the posterior predictive distribution of the missing values $y_{mis}$. The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean matching') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm $\theta$ against the iteration number. Typically, the $\theta$s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation. Monitoring convergence by inspecting traceplots of these $\theta$s is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2] **[Add that convergence of univariate thetas cannot guarantee multivariate convergence?]**. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics  may only identify signs of *non*-convergence [@hoff09]. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{mis}$. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

# Algorithmic non-convergence

@@

**[SEGWAY to this section]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$. We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $y_{imp,m}$, plotted against the iteration number. Each line portrays one imputation. The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. **In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of $j$ in $y_{imp, m}$. This difference (presumably) translates to bias in the $M$ completed data estimates $\hat{\mu}_{j,m}$, and consequently also to the pooled estimate $\bar{\mu}_j$. Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

**[MERGE with section below]**. Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with a sigle column, but the entire multivariate distribution of $y_{imp}$. Ideally, a multivariate $\theta$ should be monitored. A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. On these grounds, such a $\theta$ can be considered insufficient too. Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. Such a $\theta$ would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented. 
<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->

**[MERGE with section above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain $\theta$. $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. As @buur18 describes, researchers can specify their quantity of scientific interest $Q$ (e.g., a regression coefficient) as scalar summary $\theta$. Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.  The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a $\theta$ that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose $\lambda_1$ as such a scalar summary. We define $\lambda_1$ as the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$. $\lambda_1$ is measure that summarizes the covariances in the completed datasets. **The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by @mack03.**

@@

# Non-convergence diagnostics

@@

**[Add: we don't use effective sample size and MC error because they assume independent samples *within* chains. In MI, we only use the final estimate. Also, computing n_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand. But the default in iterative imp is often substantially lower.]**

## Potential scale reduction factor

@@

## Autocorrelation

@@

## Thresholds

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$. Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**. And finally, we will formulate an informed advice on the requirements to safely assume sufficient convergence in practice. That is, we assume that the algorithm is sufficiently converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$. **[and formulate advice on thetas?]**

@@

**[SEGWAY to this section] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $M$ completed datasets, and finally pooling the $M$ estimates $\hat{Q}$ to obtain a single $\bar{Q}$ for each $Q$. We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.**

@@

5. We further hypothesize that multivariate $\theta$s are better at detecting non-convergence than univariate $\theta$s.

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions [@vink14]. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem. Let the predictor space be defined as 
$$
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
$$
A finite population of $N=1000$ is simulated using the `mvtnorm` package [@mvtnorm]. Subsequently, a fourth variable is constructed as outcome $Y$. For each unit $i = 1, 2,..., N$, let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $\epsilon \sim \mathcal{N}(0, 100)$. From the complete set, we obtain the true values of several quantities of scientific interest, $Q$s.

@@

We consider four types of $Q$s that are often of interest in empirical research. Namely, two descriptive statistics: the mean $\mu_j$ and standard deviation $\sigma_j$ of each variable $j = Y, X_1, X_2, X_3$. We also consider the regression coefficients of the predictors: $\beta_1$, $\beta_2$, and $\beta_3$ in regression equation
$$Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,$$
where $Y'$ is the expected value of the outcome. And finally, the proportion of variance explained by the set of predictors: coefficient of determination $r^2$ **(note that lower case $r$ is used to avoid confusion with non-convergence diagnostic $\widehat{R}$)**. The true values of these scientific estimands are calculated on the complete data with the `R` functions `base::mean()`, `stats::sd()`, and `stats::lm()` [@R]. 
<!-- Each of these $Q$s is estimated with a corresponding $\bar{Q}$, after the complete dataset is first *amputed* (with a varying missingness proportion $p_{mis}$) and subsequently imputed (with a varying number of iterations $T$).  -->

@@

With these performance measures, we also evaluate how well the non-convergence diagnostics $\widehat{R}$ and $AC$ can identify simulation conditions in which the inferences are affected by non-convergence, i.e., where $\bar{Q}$ is *not* an unbiased, confidence valid estimate of $Q$. **[REPHRASE this:] For performance of univariate $Q$s, we apply the non-convergence diagnostics to univariate $\theta$s. When $Q$ is multivariate, we look at signs of non-convergence identified in multivariate $\theta$s.**

- For $Q=\mu$, $\theta$s are chain means.

- For $Q=\sigma$, $\theta$s are chain variances.

- For $Q=\beta$, $\theta$s are estimated $\beta$s per imputation.

- For $Q=r^2$, $\theta$s are the first eigenvalue of the variance-covariance matrix.

<!-- We measure the performance of $\widehat{R}$ and $AC$ by the four scalar summaries $\theta$: chain means (i.e., the mean in each $y_{imp, m, j}$) versus $Q=\mu$, chain variances (i.e., the variance in each $y_{imp, m, j}$) versus $Q=\sigma$, the estimated regression coefficients $\beta$ in each {$y_{obs}, y_{imp, m}$} (i.e., a user-defined statistic of scientific interest $\hat{Q}$) against $Q=\beta$, and the first eigenvalue of the variance-covariance matrix in each {$y_{obs}, y_{imp, m}$}, $\lambda_1$ versus $Q=r^2$.** -->

<!-- **[Remove this?] Primarily, we want to know what the effect is of inducing non-convergence on the estimated $Q$s. More specifically, we are interested in the bias of $\bar{Q}$ for all $Q$s and the confidence validity of $\bar{Q}$ for $Q=\beta$. Therefore, we evaluate these performance measures against the simulation conditions (the number of iterations $T$ and the missingness proportion $p_{mis}$). We look at averages across simulation repetitions. Secondly, we want to know if conditions in which the $Q$s are affected by non-convergence are also identified by the non-convergence diagnostics $\widehat{R}$ and $AC$. We do this, again, by looking at the performance measures for the $Q$s, but now we evaluate these against what $\widehat{R}$ and $AC$ indicate about signs of non-convergence in the $\theta$s.** -->

@@

**[Add more info about figure legends and axes]**

@@

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s. **[Add panel labels, describe panels]** Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages. Overall, we see that conditions with a higher proportion of missingness are affected more severely by non-convergence. Multivariate $Q$s are also substantially influenced by the number of iterations. This means that the algorithm has indeed not converged at $t=1$, and converges gradually as $t$ moves to $T$. The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$. The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$. This implies that, under the current specifications, no more than seven iterations are necessary to obtain sufficient performance in terms of bias and confidence validity.

### $Q=\mu$.
The estimated univariate means seem unaffected by the number of iterations in the algorithm. This implies that valid inferences may be obtained with as little as one iteration ($T \geq 1$). The magnitude of the bias in $\bar{Q}$ depends solely on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% leading to more extreme biases than other missingness proportions. Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$. Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.
